---
title: "ggplot2"
author: "John Brandt"
date: "3/10/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in the data

```{r message=FALSE, warning=FALSE}
require(ggplot2)
employment <- read.csv("data/recent-grads.csv")
employment <- na.omit(employment)
employment <- employment[-which(employment$Median > 100000),]
employment$science.other <- "Science"
employment$science.other[employment$Major_category %in%
                           c("Arts",
                             "Business",
                             "Communications & Journalism",
                             "Education",
                             "Humanities & Liberal Arts",
                             "Industrial Arts & Consumer Services",
                             "Interdisciplinary",
                             "Law & Public Policy",
                             "Psychology & Social Work",
                             "Social Science")] <- "Humanities"

#which(employment$Major=="ACTUARIAL SCIENCE")

employment$science.other[6] <- "Science"

write.csv(employment, "employment.csv")
```

## Employment data - Begin

```{r message=FALSE, warning=FALSE}
plot1 <- ggplot(data=employment, aes(x=ShareWomen, y=Median))+
  geom_point()+
  geom_smooth(se=FALSE)

print(plot1)
```

## Employment data - axis labels & theme_bw

```{r message=FALSE, warning=FALSE}
plot1 <- ggplot(data=employment, aes(x=ShareWomen, y=Median/1000))+
  geom_point()+
  geom_smooth(se=FALSE, color="grey50")+
  theme_bw()+
  xlab("Proportion of women")+
  ylab("Income (1,000 USD)")+
  ggtitle("Higher earning college majors are disproportionately men")

print(plot1)

```

## Employment data - coloring

```{r message=FALSE, warning=FALSE}
plot1 <- ggplot(data=employment, aes(x=ShareWomen, y=Median/1000))+
  geom_point(alpha=0.9, aes(color=science.other))+
  geom_smooth(se=FALSE, color="grey50")+
  theme_bw()+
  xlab("Proportion of women")+
  ylab("Income (1,000 USD)")+
  ggtitle("Higher earning college majors are disproportionately men")+
  scale_color_brewer(palette="Set1")

print(plot1)
```

## Employment data - advanced themes

```{r message=FALSE, warning=FALSE}
plot1 <- ggplot(data=employment, aes(x=ShareWomen, y=Median/1000))+
  geom_point(alpha=0.9, aes(color=science.other))+
  geom_smooth(se=FALSE, color="grey50")+
  theme_bw()+
  xlab("Proportion of women")+
  ylab("Income (1,000 USD)")+
  ggtitle("Higher earning college majors are disproportionately men")+
  scale_color_brewer(palette="Set1")+
  theme(panel.grid.minor=element_blank(),
        panel.grid.major.x=element_blank(),
        legend.title = element_blank(),
        legend.position = c(0.85, 0.8),
        legend.background=element_blank(),
        legend.text = element_text(size=10))+
  scale_x_continuous(labels=c("0%", "25%", "50%", "75%","100%"))

print(plot1)
```

## Employment data - upper + lower bounds

```{r message=FALSE, warning=FALSE}
min <- loess(employment$P25th/1000 ~ employment$ShareWomen)$fitted
max <- loess(employment$P75th/1000 ~ employment$ShareWomen)$fitted

plot1 <- ggplot(data=employment, aes(x=ShareWomen, y=Median/1000))+
  geom_point(alpha=0.9, aes(color=science.other))+
  geom_smooth(se=FALSE, color="grey50")+
  geom_smooth(aes(x=ShareWomen, y=P25th/1000), se=FALSE, color="grey85")+
  geom_smooth(aes(x=ShareWomen, y=P75th/1000), se=FALSE, color="grey85")+
  geom_ribbon(aes(ymin=min, ymax=max), alpha=0.08)+
  theme_bw()+
  xlab("Proportion of women")+
  ylab("Income (1,000 USD)")+
  ggtitle("Higher earning college majors are disproportionately men")+
  scale_color_brewer(palette="Set1")+
  theme(panel.grid.minor=element_blank(),
        panel.grid.major.x=element_blank(),
        legend.title = element_blank(),
        legend.position = c(0.85, 0.8),
        legend.background=element_blank(),
        legend.text = element_text(size=10))+
  scale_x_continuous(labels=c("0%", "25%", "50%", "75%","100%"))

print(plot1)
```


## Employment data (bar) - begin

```{r}
plot2 <- ggplot(data=employment, aes(x=Major_category, y=mean(Median)))+
  geom_col()

print(plot2)
```

## Employment data (bar) - restructure data

```{r message=FALSE, warning=FALSE}
require(dplyr)

employment.sub <- employment[employment$Sample_size >= 50,]

top <- employment.sub %>%
  arrange(desc(Median))
top <- top[0:10,]

bottom <- employment.sub %>%
  arrange(Median)
bottom <- bottom[0:10,]

topbottom <- rbind(top, bottom)

topbottom$wage.diff <- topbottom$Median - mean(employment$Median)

head(topbottom[,c(3,16)])
```

## Plot 2 - Simple plot + coord_flip + themes + labels

```{r}
plot2.f <- ggplot(data=topbottom, aes(x=reorder(Major, Median), y=Median/1000))+
  geom_col()+
  coord_flip()+
  theme_bw()+
  xlab("")+
  ylab("Earnings (1,000 USD)")+
  ggtitle("Earnings of top and bottom ten majors")

print(plot2.f)
```

## Plot 2 - Coloring + advanced themes

```{r}
plot2.f <- ggplot(data=topbottom, aes(x=reorder(Major, Median), y=Median/1000))+
  geom_col(aes(fill=science.other))+
  coord_flip()+
  theme_bw()+
  xlab("")+
  ylab("Earnings (1,000 USD)")+
  ggtitle("Earnings of top and bottom ten majors")+
  theme(legend.position=c(0.8, 0.2),
        legend.background=element_rect(fill="transparent"),
        legend.title=element_blank(),
        legend.text = element_text(size=10),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank())

print(plot2.f)
```

## Plot 2 - Mean line + annotation

```{r}
plot2.f <- ggplot(data=topbottom, aes(x=reorder(Major, Median), y=Median/1000))+
  geom_col(aes(fill=science.other))+
  geom_hline(aes(yintercept=mean(employment$Median/1000)), linetype="dashed", color="grey10")+
  scale_fill_brewer(palette="Set1")+
  coord_flip()+
  theme_bw()+
  theme(legend.position=c(0.8, 0.2),
        legend.background=element_rect(fill="transparent"),
        legend.title=element_blank(),
        legend.text = element_text(size=10),
        axis.ticks.y = element_blank(),
        panel.grid.major.y = element_blank())+
  annotate("text", x=1, y=50, label="Median wage")+
  xlab("")+
  ylab("Earnings (1,000 USD)")+
  ggtitle("Earnings of top and bottom ten majors")

print(plot2.f)
```


## Weather - begin

```{r}
los.angeles <- read.csv("data/KCQT.csv")
los.angeles$date <- as.Date(los.angeles$date)
```

```{r}
plot3 <- ggplot(data=los.angeles, aes(x=date, y=actual_mean_temp))+
  geom_line()

print(plot3)
```


## Weather - theme_bw + labels

```{r}
plot3 <- ggplot(data=los.angeles, aes(x=date, y=actual_mean_temp))+
  geom_line()+
  theme_bw()+
  xlab("")+
  ylab("Temperature (F)")+
  ggtitle("2014 Weather in Los Angeles, CA")
  

print(plot3)
```

## Weather - units

```{r}
plot3 <- ggplot(data=los.angeles, aes(x=date, y=actual_mean_temp))+
  geom_line()+
  theme_bw()+
  xlab("")+
  ylab("Temperature (F)")+
  ggtitle("2014 Weather in Los Angeles, CA")+
  scale_y_continuous(breaks=c(seq(20,110, 10)))+
  scale_x_date(expand=c(0, 0), date_breaks="1 month", date_labels = "%b")
  

print(plot3)
```

## Weather - overlay average + record

```{r}
plot3 <- ggplot(data=los.angeles, aes(x=date, y=actual_mean_temp))+
  theme_bw()+
  geom_linerange(aes(ymin=record_min_temp, ymax=record_max_temp, color="Record"), size=1.5)+
  geom_linerange(aes(ymin=actual_min_temp, ymax=actual_max_temp, color="Actual"), size=1.5)+
  geom_linerange(aes(ymin=average_min_temp, ymax=average_max_temp, color="Average"), size=1.5, alpha=0.2) +
  geom_point(size=0.7, color="grey30")+
  scale_y_continuous(breaks=c(seq(20,110, 10)))+
  scale_x_date(expand=c(0, 0), date_breaks="1 month", date_labels = "%b") + 
  xlab("")+
  ylab("Temperature (F)")+
  ggtitle("2014 Weather in Los Angeles, CA")

print(plot3)
```


## Weather - colors + themes

```{r}
plot3.f <- ggplot(data=los.angeles, aes(x=date, y=actual_mean_temp))+
  theme_bw()+
  geom_linerange(aes(ymin=record_min_temp, ymax=record_max_temp, color="Record"), size=1.5)+
  geom_linerange(aes(ymin=actual_min_temp, ymax=actual_max_temp, color="Actual"), size=1.5)+
  geom_linerange(aes(ymin=average_min_temp, ymax=average_max_temp, color="Average"), size=1.5, alpha=0.2) +
  geom_point(size=0.7, color="grey30")+
  geom_point(data=subset(los.angeles, actual_max_temp >= record_max_temp), aes(y=actual_max_temp), color="red", size = 0.8)+
  scale_colour_manual(labels=c("Actual", "Average", "Record"), values=c("grey70", "grey60", "bisque2"))+
  theme(panel.grid.minor=element_blank(),
        panel.grid.major=element_line(linetype="dashed"),
        legend.title=element_blank(),
        legend.direction = "horizontal",
        legend.position = c(0.18,0.05),
        legend.background=element_rect(colour="transparent", fill="transparent"),
        panel.background = element_rect(fill = NA))+
  scale_y_continuous(breaks=c(seq(20,110, 10)))+
  scale_x_date(expand=c(0, 0), date_breaks="1 month", date_labels = "%b") + 
  xlab("")+
  ylab("Temperature (F)")+
  ggtitle("2014 Weather in Los Angeles, CA")+
  annotate("text", x=as.Date("2014-10-06"), y=los.angeles$actual_max_temp[which.max(los.angeles$actual_max_temp)] + 0.5, label="- Record", color="red")

print(plot3.f)
```

## Weather data - advanced

```{r}
plot3.f <- ggplot(data=los.angeles, aes(x=date, y=actual_mean_temp))+
  theme_bw()+
  geom_linerange(aes(ymin=record_min_temp, ymax=record_max_temp, color="Record"), size=1.5)+
  geom_linerange(aes(ymin=actual_min_temp, ymax=actual_max_temp, color="Actual"), size=1.5)+
  geom_linerange(aes(ymin=average_min_temp, ymax=average_max_temp, color="Average"), size=1.5, alpha=0) +
  geom_ribbon(aes(ymin=average_min_temp, ymax=average_max_temp), color="transparent", fill="grey50", size=0.5, alpha=0.5) +
  geom_point(size=0.7, color="grey30")+
  scale_colour_manual(name="x", labels=c("Actual", "Average", "Record"), values=c("grey70", "grey90", "bisque2"))+
  theme(panel.grid.minor=element_blank(),
        panel.grid.major=element_line(linetype="dashed"),
        legend.title=element_blank(),
        legend.direction = "horizontal",
        legend.position = c(0.18,0.05),
        legend.background=element_rect(colour="transparent", fill="transparent"))+
  scale_y_continuous(breaks=c(20,30,40,50,60,70,80,90,100,110,120))+
  scale_x_date(expand=c(0, 0), date_breaks="1 month", date_labels = "%b") + 
  xlab("")+
  ylab("Temperature (F)")+
  ggtitle("2014 Weather in Los Angeles, CA")

print(plot3.f)
```

