---
title: "Rshiny"
author: "John Brandt"
date: "3/15/2018"
output: beamer_presentation
---

```{r setup, include= FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=FALSE)
```

## Packages & Data

```{r}
library(shiny)
library(ggvis)
library(tibble)
library(magittr)

data <- mtcars
colnames(data) <- c("MPG", "Number_cylinders", 
      "Displacement", "Horsepower", 
      "Rear_axle_ratio", "Weight_1000_lb", 
      "Quarter_m_time", "Engine_shape", "Transmission",
      "Forward_gears", "Carburetors")
```

## Process data

```{r}
data$Transmission[data$Transmission == 0 ] <- "Automatic"
data$Transmission[data$Transmission == 1] <- "Manual"
data$Transmission <- factor(data$Transmission) 

data$Engine_shape[data$Engine_shape == 0] <- "V-shape"
data$Engine_shape[data$Engine_shape == 1] <- "S-shape"
data$Engine_shape <- factor(data$Engine_shape)

data$Forward_gears <- factor(data$Forward_gears)
data$Carburetors <- factor(data$Carburetors)

data$Number_cylinders <- factor(data$Number_cylinders)
```

## Subset data for varible selections

```{r}
axis_vars <- c("MPG", "Displacement", "Horsepower", 
               "Rear_axle_ratio", "Weight_1000_lb",
               "Quarter_m_time")

factor_vars <- c("Number_cylinders", "Transmission", 
                 "Forward_gears", "Carburetors", 
                 "Engine_shape")

data <- tibble::rownames_to_column(data, var="Model")
```

## Rshiny basics

* UI & Server
* fluidRow - rows that expand to monitor/screen size
* Columns, relative sizing
* https://shiny.rstudio.com/gallery/widget-gallery.html
    + Many widgets
    + selectInput
    + sliderInput

## selectInput

```{r}
selectInput("reference.name", "label", choices, 
            selected = "Default")
```

Now, in the server, `input$reference.name` will interactively update to the input

## Create UI

```{r}
ui <- fluidPage(
  titlePanel("Yale NUS"),
  
  fluidRow(
    column(3,
           ## Next Slide
    ),
```

## Input 
```{r}
helpText("Explore relationships between 
    fuel efficiency and vehicle properties"),
selectInput("yvar", "Y-axis variable", axis_vars, 
    selected = "Horsepower"),
selectInput("xvar", "X-axis variable", axis_vars,
    selected = "MPG"),
selectInput("loess", "Linear model variable", 
            factor_vars, 
    selected = "Number_cylinders"),
sliderInput("range", label= "Horsepower:",min = 0, 
            max = 350, value = c(0,350))
```

## Column (75% of space) with visualization

```{r}
    column(9,
           ggvisOutput("plot1")
    )
  )
)
```

## Create server

Server must have input, output

```{r}
# Define server logic ----
server <- function(input, output) {
```

## Create interactive tooltip

```{r}
  car_tooltip <- function(x) {
    if (is.null(x)) return(NULL)
    # Pick out the car with this model
    column <- yvar$value
    row <- xvar$value
    if(is.null(x[[column]])) return(NULL)
    car <- data[data[[column]] == x[[column]] & 
                  data[[row]] == x[[row]],]
    paste0("<b>", as.character(car$Model), "</b><br>")
  }
```

## Add reactivity to visualization
```{r}
  vis <- reactive({
    # Labels for axes
    xvar_name <- names(axis_vars)[axis_vars == input$xvar]
    yvar_name <- names(axis_vars)[axis_vars == input$yvar]
    xvar <<- prop("x", as.symbol(input$xvar))
    yvar <<- prop("y", as.symbol(input$yvar))
    factorvar <<- prop("factor", as.symbol(input$loess))
```

## Update the dataframe used & create ggvis
```{r}
    data <- data[data$Horsepower >= input$range[1],]
    data <- data[data$Horsepower <= input$range[2],]
    
    data %>%
      group_by(data[[factorvar$value]]) %>%
      ggvis(xvar, yvar) %>%
      layer_model_predictions(stroke= 
                                factorvar$value, 
                              model="lm") %>%
      layer_points(fill = factorvar$value) %>%
      add_tooltip(car_tooltip, on="hover") %>%
      add_axis("x", title = xvar_name) %>%
      add_axis("y", title = yvar_name) %>%
      set_options(width = 1000, height = 700) %>%
      add_legend("stroke")
  })
```

## Add the visualization

```{r}
  vis %>% bind_shiny("plot1")
}
```

## Run the app

```{r}
# Run the app ----
shinyApp(ui = ui, server = server, options = 
           list(height= 700))
```

